### 2015-05-19 Manuscript/Missing_mammals-draft.tex

%Pagination style and stuff
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

@@

\noindent{\Large \bf Morphological data availability in living mammals}
%Or: Missing cladistic data in living mammals - but I prefer the first one emphasizing on the availability of it. The data is not missing per se, it's just not coded!

@@

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$$^*$, and Natalie Cooper$^1$$^,$$^2$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\
\vspace{1in}

%Line numbering
%\modulolinenumbers[1]
%\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------

\newpage
\begin{abstract}
Studying changes in global biodiversity through time and space is essential.
For that we need methods to combine both palaeontological and neontological data.
One promising method, the Total Evidence method, allows such thing but needs a lot of data.
Especially cladistic data from living taxa to allow the fossil taxa to accurately branch in the trees.
Despite two centuries of morphological studies on living taxa, scientists using and generating such data mainly focus on palaeontological data.
Therefore, even in well known groups such as mammal, there is a huge gap in our knowledge of cladistic data for living mammals.
In this study, using phylogenetic community structure methods, we quantify the availability of data in each mammalian order.
And maybe at the end of the paper we propose some discussion on how to improve all that (go in museums!).
\end{abstract}

\noindent ()\\

\vspace{1.5in}

\newpage

@@

\section{Introduction}
Studying both living and fossil taxa together in essential to fully understand macrovelutionary patterns and processes and is becoming increasingly common among evolutionary biologists \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013,Wood01032013}.
Combining the global biodiversity by including both living and fossil in such studies allows, for example, to improve accuracy of the timing of diversification events \citep[e.g.][]{ronquista2012}, to better understand relationships among lineages \citep[e.g.][]{beckancient2014} or to infer biogeographical patterns through time \citep[e.g.][]{Meseguer01032015}.
%And add STD paper? Or that's too much self cites?
In order to perform such analysis, one must efficiently combine living and fossil taxa data in macroevolutionary models.
One trending method, called the Total Evidence method \citep{eernissetaxonomic1993,ronquista2012}, allows to combine molecular data from living taxa and cladistic data from both living and fossil taxa in a supermatrix \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014,Meseguer01032015}.
This method not only allows to use all the available data but also allows to treat fossil taxa as tips rather than nodes \textit{via} integrative phylogenetic inference methods such as tip-dating \citep{ronquista2012,Drummond01082012,Wood01032013,BEASTmaster}.

However, the Total Evidence method requires, by definition, a lot of data.
One must collect both molecular data for living taxa as well as cladistic data for both living and fossil taxa, two types of data that require fairly different technical skills \citep[e.g.][]{meredithimpacts2011} \textit{vs.} \citep{O'Leary08022013}.
Additionally, entire sections of this data can sometimes be difficult or impossible to collect for every taxa present in the analysis.
For example, fossil have really rarely molecular data available and morphological characters are rarely collected for living taxa when molecular data is available \citep[e.g.][]{slaterphylogenetic2013,beckancient2014}.
This difference between both data can lead to topological errors in phylogenetic inference \citep{GuillermeCooper}.
In fact, the ability of the Total Evidence method to recover correct topology is expected to decrease when there is a low overlap between cladistic data from the living and the fossil taxa \citep{GuillermeCooper}.
The effect of missing data is most important on topology when too few living taxa have available cladistic data \citep{GuillermeCooper}.
For example, if there is no morphological available for any living taxa within an entire clade, it is impossible to link a fossil taxon to this clade because no cladistic data will overlap between the fossil (regardless the amount of data available for the fossil) and the living taxa (that have no cladistic data).
This property of the Total Evidence method can rapidly lead to important topological incongruities because the fossil will only be able to branch to a clade of living taxa that contains cladistic data, even if in reality, the fossil taxon does not belong to that clade \citep{GuillermeCooper} (see figure ~\ref{Figure_missing_data_problem}).

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{MissingDataFigure.pdf}
\caption{Example of topological errors due to missing cladistic data in living taxa.
If a phylogeny contains two clades, for example Aves and Mammalia, with molecular data (in grey) for both but only cladistic data (in red) for Aves.
If an additional Mammalia fossil (with no molecular data) is added to the phylogeny, it will erroneously branch with the Aves clade instead of the Mammalia one because no cladistic data will overlap between the fossil Mammalia and the living ones.}
\label{Figure_missing_data_problem}
\end{figure}

%What we're doing
It is therefore crucial to understand the distribution of the available cladistic data for living taxa in a clade before using a Total Evidence approach because missing living taxa with cladistic data can lead to two topological artefacts: (1) the impossibility for a fossil to branch in the right clade if there is no cladistic data available in this clade \citep{GuillermeCooper}; and (2) the higher probability for a fossil to branch within a wrong clade that has more cladistic data available for living taxa than the right clade \citep{GuillermeCooper}.
In this study, we assess the level of available cladistic data in living mammals in order to highlight the two caveats described above.
%In this study we investigate the amount of living mammal taxa with available cladistic data to detect the two potential caveats described above that could lead to topological errors when building Total Evidence mammal phylogenetic trees.
We collected living mammals cladistic data from @256 phylogenetic matrices available online and measured the proportion of cladistic data availability for each mammalian order.
Additionally, in the mammalian orders where data was missing, we estimated the structure of the available data to detect if the available data was biased towards certain clades only or if it was randomly distributed using the Net Relatedness Index \citep{webb2002phylogenies}.

>>>>

%Pagination style and stuff % NC: Note that these are all syst biol specific.
\linespread{2}
\raggedright
\setlength{\parindent}{0.5in}
\setcounter{secnumdepth}{0} 
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
%\bibpunct{(}{)}{;}{a}{}{,}

@@

\noindent{\Large \bf Morphological data availability in living mammals}
%Or: Missing cladistic data in living mammals - but I prefer the first one emphasizing on the availability of it. The data is not missing per se, it's just not coded!
% NC: sometimes the better title is one that answers the question rather than giving the topic. e.g., "Morphological data is lacking for living species" or similar

@@

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$$^,$$^*$ and Natalie Cooper$^1$$^,$$^2$$^,$$^3$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
$^3$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\ % NC: Might need to reconsider the email here as it won't be your email anymore when you leave. Or do you get to keep it forever? 
\vspace{1in}

%Line numbering
\modulolinenumbers[1]
\linenumbers

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
% NC: I've commented this out so I don't have to keep scrolling past in in the PDF
%\newpage
%\begin{abstract}
%Studying changes in global biodiversity through time and space is essential.
%For that we need methods to combine both palaeontological and neontological data.
%One promising method, the Total Evidence method, allows such thing but needs a lot of data.
%Especially cladistic data from living taxa to allow the fossil taxa to accurately branch in the trees.
%Despite two centuries of morphological studies on living taxa, scientists using and generating such data mainly focus on palaeontological data.
%Therefore, even in well known groups such as mammal, there is a huge gap in our knowledge of cladistic data for living mammals.
%In this study, using phylogenetic community structure methods, we quantify the availability of data in each mammalian order.
%And maybe at the end of the paper we propose some discussion on how to improve all that (go in museums!).
%\end{abstract}

%\noindent ()\\

%\vspace{1.5in}

@@

%NC: I've changed cladistic to morphological throughout but you might want to change it back...

\newpage 
\section{Introduction}
There is an increasing consensus among evolutionary biologists that studying both living and fossil taxa is essential for fully understanding macroevolutionary patterns and processes \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013,Wood01032013}.
For example, including both living and fossil taxa in evolutionary studies can improve the accuracy of timing diversification events \citep[e.g.][]{ronquista2012}, our understanding of relationships among lineages \citep[e.g.][]{beckancient2014}, and our ability to infer biogeographical patterns through time \citep[e.g.][]{Meseguer01032015}.
To perform such analyses it is necessary to combine living and fossil taxa in phylogenetic trees.
One increasingly popular method, the Total Evidence method \citep{eernissetaxonomic1993,ronquista2012}, combines molecular data from living taxa and morphological data from both living and fossil taxa in a supermatrix \citep[e.g.][]{pyrondivergence2011,ronquista2012,schragocombining2013,slaterunifying2013,beckancient2014,Meseguer01032015}, producing a phylogeny with living and fossil taxa at the tips. 
These phylogenies can be dated using methods such as tip-dating \citep{ronquista2012,Drummond01082012,Wood01032013,BEASTmaster} and incorporated into macroevolutionary studies (e.g. ).

A downside of the Total Evidence method is that it requires a lot of data.
One must collect molecular data for living taxa and morphological data for both living and fossil taxa; two types of data that require fairly different technical skills \citep[e.g.][]{meredithimpacts2011} \textit{vs.} \citep{O'Leary08022013}.
% NC: Is that really what these papers show? They use differnt data but they don't talk about the technical skills! Might be better just to say something like (e.g., PCR/sequencing vs. anatomical description)
Additionally, large chunks of this data can be difficult, or even impossible, to collect for every taxon present in the analysis.
For example, fossils very rarely have molecular data and incomplete fossil preservation may restrict the amount of morphological data available. % NC: Worth metioning fossil preservation here briefly as it is also an issue?
Additionally, since the molecular phylogenetics revolution, it has become less common to collect morphological characters for living taxa when molecular data is available \citep[e.g.][]{slaterphylogenetic2013,beckancient2014}.
%NC: Do these references actually say this??? Sounds like that's what you are saying
Unfortunately this missing data can lead to errors in phylogenetic inference; in fact, simulations show that the ability of the Total Evidence method to recover the correct phylogenetic topology decreases when there is a low overlap between morphological data in the living and fossil taxa \citep{GuillermeCooper}, regardless the overall amount of morphological data available for the fossils (or the amount of molecular data available for the living species).
The effect of missing data on topology is greatest when living taxa have little morphological data.
This is because (1) a fossil cannot branch in the correct clade if there is no overlapping morphological data in the clade; and (2) a fossil has a higher probability of branching within a clade with more morphological data available for living taxa, regardless of whether this is the correct clade or not \citep{GuillermeCooper}. 
This can lead to serious topological errors (see Figure ~\ref{Figure_missing_data_problem}).

% NC: OK so I know we did a paper on this but you can't just cite it every sentence. Need to put in a few more refs to other people's work or just cite us once or twice. I've condensed it which should help

\begin{figure}[!htbp]
\centering
    \includegraphics[width=1\textwidth]{MissingDataFigure.pdf}
\caption{Example of topological errors due to missing morphological data in living taxa.
The phylogeny contains two clades, Aves and Mammalia, with molecular data (grey) for both but only morphological data (red) for Aves.
If a mammalian fossil species (with no molecular data) is added to the phylogeny, it will erroneously branch with the Aves instead of the Mammalia because no morphological data will overlap between the fossil mammals and the living ones.}
\label{Figure_missing_data_problem}
\end{figure}

%What we're doing

The issues above highlight that it is crucial to have sufficient morphological data for living taxa in a clade before using a Total Evidence approach.
However, it is unclear how much morphological data for living taxa is actually available (i.e. already coded from specimens of living species and deposited in phylogenetic matrices), and how this data is distributed.
Intuitively, most people assume this kind of data has already been collected, but our experiences suggest otherwise.
To investigate this further, here we assess the amount of available morphological data for living mammals to determine whether sufficient data exists to build reliable Total Evidence phylogenies in this group.
We collected cladistic data from @256 phylogenetic matrices available online and measured the proportion of cladistic data available for each mammalian order.
Additionally, if the available data is randomly distributed across clades, the topological errors should be less extreme than if the data is biased towards particulars groups.
Therefore, in the mammalian orders where data was missing, we determined whether the available cladistic data was phylogenetically overdispersed or clustered. 
We find???
% NC: Needs some summing up sentence or a we find sentence?
